DNA Methylation Hierarchical Variance Model\nPaper Figures

### 2021-07-07 Paper_Figures.Rmd
updates to figures and code

stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
load(here(stanFile))
CpGscore_full = CpGscore_full %>% mutate(mean = mean*2, median=median*2)

prepdatafolder = "data_final_rg3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed in StanAnalysis.Rmd)...",prepdatafolder))
gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))

gened_data <- gened_data %>% mutate(score_median = score_median*2)
prepped_data <- prepped_data %>% mutate(score_median = score_median*2)
permed_data <- permed_data %>% mutate(score_median = score_median*2)

@@

$$\text{score}_\text{med} = \text{Median} (\text{log}_2\frac{\sigma_{P}}{\sigma_T}) $$

\newpage
\subsection{Figure 2A: Example site fit, 2B: Posteriors}
![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_fit.png)
\newline \newline
![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_posteriors.png)

# plot hierarchical model posterior medians for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
pM <- post_prior_histogram(mu_full$p50, function(x) 10*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
pN <- post_prior_histogram(nu_full$p50, function(x) 15*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
#pE <- post_prior_histogram(sigmaE_full$p50, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
#pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
#pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
#pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
pE <- post_prior_histogram(sigmaE_full$p50^2, function(x) 100*dgamma(sqrt(x),2,2), 0.05, c(0,2), "SigmaE^2")
pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 250*dgamma(sqrt(x), 2.7693, 7.0029), 0.01, c(0,1), "SigmaP^2")
pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 40*dgamma(sqrt(x), 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT^2")
pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 220*dgamma(sqrt(x), 3.3643, 10.430), 0.01, c(0,1), "SigmaT^2")
# log sigmas
#pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
#pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
#pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))

@@

kable(Rhat_mat, col.names=c("Parameter", "Mean Rhat", "Max Rhat", "# >1.1", "% >1.1", "# >1.01", "% >1.01"), digits = 2)
cat("  \n")

# indices to filter out sites Rhat>1,1
QCinds <- lp_full$Rhat > 1.1
```

For the remainder of these results, a filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit). Another filter is later applied to remove all genes with <5 CpG sites.

@@

# overlaid transparent histograms
post_histogram <- function(postvals,mask,bw,titlestr) {
  # transparent histogram colored by gene-associated sites
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  gene_assoc_df <- plot_df %>% filter(gene_assoc)
  not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
    geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
    ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
  p1
}

# plot the two conservations score prob sums
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
p_scores <- post_histogram(CpGscore_full$median,genemask,0.2,"CpG Conservation score") +
  theme(aspect.ratio=1/4,legend.position = "none") +
  scale_x_continuous(limits=c(-8,8),breaks=c(-8,-4,0,4,8)) + 
  scale_y_continuous(breaks=c(0,0.1,0.2,0.3,0.4,0.5))
print(p_scores)

df = data.frame(med=CpGscore_full$median, mask=genemask)
p_scores <- ggplot(data=df) + geom_boxplot(aes(x=med, y = genemask, fill = genemask), outlier.shape = NA) +
  scale_x_continuous(limits=c(-8,8),breaks=c(-8,-4,0,4,8)) + xlab("CpG score") + ylab("Gene-associated") + theme_classic() + theme(aspect.ratio=1/4,legend.position = "none")
print(p_scores)
# ggsave("CpGscores_overlay.svg")

# Fig 4B: plot gene scores and adjusted bootstrap distribution
p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-3,7)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p_genes <- ggarrange(p1,p2,nrows=1,ncol=2)
# ggsave("gene_scores_bootstrap.svg")
```

@@

# Fig 3C: plot functional region score distributions
temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
  separate_rows(UCSC_RefGene_Group, sep=";") %>%
  distinct() %>%
  dplyr::rename(fr = UCSC_RefGene_Group, score=score_median) %>%
  mutate(fr = case_when(fr=="" ~ "NA",
                        str_detect(fr, "5URT") ~ "5'UTR",
                        str_detect(fr, "3UTR") ~ "3'UTR",
                        TRUE ~ fr)) %>%
  distinct()
#regions = c("TSS1500","Body", "3'UTR","1stExon","TSS200","5'UTR","ExonBnd")
regions = rev(c("TSS200","1stExon", "5'UTR","TSS1500","Body","3'UTR","ExonBnd"))
temp %>%
  mutate(fr = factor(fr, levels = regions)) %>%
  filter(fr %in% regions) %>%
  #ggplot() + geom_violin(aes(x=fr,y=score,fill=fr), draw_quantiles = 0.5) + theme_classic() +
  ggplot() + geom_boxplot(aes(y=fr,x=score, fill=fr), outlier.shape = NA) + theme_classic() +
  scale_x_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none") + geom_vline(xintercept = 0, color="black", alpha=0.2)
# ggsave("CpGscore_regions_violin.svg")
ggsave("CpGscore_regions_boxplot.svg")
#rm(temp)3

@@

# plot for conservation scores
distance_length %>% drop_na() %>%
  ggplot(aes(x = location_bin, y = mean_score_median)) +
  geom_point(color="black") +
  theme_bw() + theme(legend.position = "none", aspect.ratio = 3/8) +
  xlab("Distance from gene start site (binned to 100 bp)") + ylab("Average Conservation Score") +
  scale_x_continuous(breaks = seq(-1500,10000,1500))
cat("\n")
# ggsave("CpGdist.svg")
```

@@

cat("\n")
# Fig 4A: Example genes CpG site distributions
plotGeneScore("TUBA1A",0.2)
#ggsave("genescore_TUBA1A.svg")
cat("\n")
plotGeneScore("TTN",0.26)
#ggsave("genescore_TTN.svg")
cat("\n")
plotGeneScore("HLA-A",0.18)
#ggsave("genescore_HLA-A.svg")

@@

# CRC gene database 1: COSMIC
cancergenes1 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
# CRC gene database 2: AGO
cancergenes2 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")

@@

# plot violins of: all genes, COSMIC, AGO
perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n))
genes_dbs <- rbind(perm_temp %>% mutate(group="all"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% mutate(group="COSMIC"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO")
              )
# Fig 4C: plot distribution
pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all", "AGO", "COSMIC"))) %>%
  #ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) +
  ggplot() + geom_boxplot(aes(x=score_median,y=group,fill=group), outlier.shape = NA) +
  scale_x_continuous(limits=c(-2,4),breaks = c(-2,0,2,4)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
# # add hashmarks for each gene (doesn't look great)
# marks <- genes_dbs %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
# marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
# pDB + geom_segment(data=marks2, aes(x=x, xend=x, y=y, yend=y+0.2), color="red") + geom_segment(data=marks, aes(x=x, xend=x, y=y, yend=y+0.1), color="black", alpha=0.5)
pDB
# ggsave("database_violins.svg")
ggsave("database_boxplot.svg")

# stats
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median[genes_dbs$group=="all"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="AGO"], y = genes_dbs$score_median[genes_dbs$group=="all"], alternative = "greater")
```

>>>>

stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
load(here(stanFile))

prepdatafolder = "data_final_rg3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed in StanAnalysis.Rmd)...",prepdatafolder))
gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))
```

@@

\newpage
\subsection{Figure 2A: Example site fit, 2B: Example site posteriors}
![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_fit.png)
\newline \newline
![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_posteriors.png)

@@

kable(Rhat_mat, col.names=c("Parameter", "Mean Rhat", "Max Rhat", "# >1.1", "% >1.1", "# >1.01", "% >1.01"), digits = 2)
cat("  \n")
```

\subsection{Filter CpG sites}
A filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit), and CpG sites mapping to sex chromosomes X and Y are removed as well to eliminate sex-specific bias.

Later, a gene filter is applied to remove all genes with <5 CpG sites.

```{r filter_sites, echo=FALSE, warning=FALSE}
# indices to filter out sites Rhat>1,1
QCinds <- which(lp_full$Rhat > 1.1)
# indices to filter out X and Y chromosome
XYinds <- which(FullAnnotation$CHR %in% c("X","Y"))

# remove the corresponding rows from the results dataframes
FullAnnotation <- FullAnnotation[-union(QCinds,XYinds),]
mu_full <- mu_full[-union(QCinds,XYinds),]
nu_full <- nu_full[-union(QCinds,XYinds),]
sigmaP_full <- sigmaP_full[-union(QCinds,XYinds),]
sigmaPT_full <- sigmaPT_full[-union(QCinds,XYinds),]
sigmaT_full <- sigmaT_full[-union(QCinds,XYinds),]
sigmaE_full <- sigmaE_full[-union(QCinds,XYinds),]
lp_full <- lp_full[-union(QCinds,XYinds),]
CpGscore_full <- CpGscore_full[-union(QCinds,XYinds),]
```

\subsection{Figure 2C: Posterior medians over all CpG sites}
```{r posteriors, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
# for each parameter, plot the distribution and overlay the fits
fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
post_prior_histogram <- function(postvals,priorfun,bw,xl,titlestr) {
  # histogram with prior curve overlaid
  plot_df <- data.frame(val=postvals)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..density..), binwidth=bw, position="dodge", fill="lightgray") +
    stat_function(fun=function(x) priorfun(x), geom="line", color="black") +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
  p1
}

# plot mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_full$p50, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu") + theme(aspect.ratio = 1)
pN <- post_prior_histogram(nu_full$p50, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "Nu") + theme(aspect.ratio = 1)
pE <- post_prior_histogram(sigmaE_full$p50, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE") + theme(aspect.ratio = 1)
pP <- post_prior_histogram(sigmaP_full$p50, function(x) 1.52*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP") + theme(aspect.ratio = 1)
pPT <- post_prior_histogram(sigmaPT_full$p50, function(x) 1.6*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT") + theme(aspect.ratio = 1)
pT <- post_prior_histogram(sigmaT_full$p50, function(x) 0.97*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT") + theme(aspect.ratio = 1)

p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
print(p_all)
rm(pM,pN,pE,pP,pPT,pT)

ggsave(filename = "posteriors.svg", plot=p_all)
```

# plot hierarchical model posterior medians for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
pM <- post_prior_histogram(mu_full$p50, function(x) 10*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
pN <- post_prior_histogram(nu_full$p50, function(x) 15*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
#pE <- post_prior_histogram(sigmaE_full$p50, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
#pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
#pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
#pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
pE <- post_prior_histogram(sigmaE_full$p50^2, function(x) 100*dgamma(sqrt(x),2,2), 0.05, c(0,2), "SigmaE^2")
pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 250*dgamma(sqrt(x), 2.7693, 7.0029), 0.01, c(0,1), "SigmaP^2")
pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 40*dgamma(sqrt(x), 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT^2")
pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 220*dgamma(sqrt(x), 3.3643, 10.430), 0.01, c(0,1), "SigmaT^2")

$$\text{score}_\text{med} = \text{Median} (\text{log}_2\frac{\sigma_P^2}{\sigma_T^2}) $$

@@

# overlaid transparent histograms
# post_histogram <- function(postvals,mask,bw,titlestr) {
#   # transparent histogram colored by gene-associated sites
#   plot_df <- data.frame(val=postvals, gene_assoc=mask)
#   gene_assoc_df <- plot_df %>% filter(gene_assoc)
#   not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
#   p1 <- ggplot(plot_df, aes(x=val)) +
#     geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
#     geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
#     ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
#   p1
# }

# plot conservation score
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
p_scores <- post_histogram(CpGscore_full$median,genemask,0.2,"CpG Conservation score") +
  theme(aspect.ratio=1/4,legend.position = "none") +
  scale_x_continuous(limits=c(-8,8),breaks=c(-8,-4,0,4,8)) + 
  scale_y_continuous(breaks=c(0,0.1,0.2,0.3,0.4,0.5))
print(p_scores)
ggsave("CpGscores_dodge.svg")

# same data but as boxplot
df = data.frame(med=CpGscore_full$median, mask=genemask)
p_scores <- ggplot(data=df) + geom_boxplot(aes(x=med, y = genemask, fill = genemask), outlier.shape = NA) +
  scale_x_continuous(limits=c(-8,8),breaks=c(-8,-4,0,4,8)) + xlab("CpG score") + ylab("Gene-associated") + theme_classic() + theme(aspect.ratio=1/4,legend.position = "none")
print(p_scores)

# statistics: two-sample wilcoxon rank-sum
wilcox.test(x=CpGscore_full$median[genemask], y=CpGscore_full$median[!genemask], alternative = "greater")

# Fig 4B: plot gene scores and adjusted bootstrap distribution
p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.2, fill="lightgray",color = "grey50") + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1) + scale_x_continuous(breaks = c(-2,0,2,4,6), limits = c(-3,7)) + scale_y_continuous(breaks = c(0,1000,2000,3000), limits = c(0,3000))
p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p_genes <- ggarrange(p1,p2,nrows=1,ncol=2)
ggsave("gene_scores_median.svg", plot=p1)
ggsave("gene_scores_bootstrap.svg", plot=p2)
```

@@

# Fig 3C: plot functional region score distributions
temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
  separate_rows(UCSC_RefGene_Group, sep=";") %>%
  distinct() %>%
  dplyr::rename(fr = UCSC_RefGene_Group, score=score_median) %>%
  mutate(fr = case_when(fr=="" ~ "NA",
                        str_detect(fr, "5URT") ~ "5'UTR",
                        str_detect(fr, "3UTR") ~ "3'UTR",
                        TRUE ~ fr)) %>%
  distinct()
#regions = c("TSS1500","Body", "3'UTR","1stExon","TSS200","5'UTR","ExonBnd")
regions = rev(c("TSS200","1stExon", "5'UTR","TSS1500","Body","3'UTR","ExonBnd"))
temp %>%
  mutate(fr = factor(fr, levels = regions)) %>%
  filter(fr %in% regions) %>%
  #ggplot() + geom_violin(aes(x=fr,y=score,fill=fr), draw_quantiles = 0.5) + theme_classic() +
  ggplot() + geom_boxplot(aes(y=fr,x=score, fill=fr), outlier.shape = NA) + theme_classic() +
  scale_x_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none") + geom_vline(xintercept = 0, color="black", alpha=0.2)
# ggsave("CpGscore_regions_violin.svg")
ggsave("CpGscore_regions_boxplot.svg")
#rm(temp)

@@

# plot for conservation scores
distance_length %>% drop_na() %>%
  ggplot(aes(x = location_bin, y = mean_score_median)) +
  geom_point(color="black") +
  theme_bw() + theme(legend.position = "none", aspect.ratio = 3/8) +
  xlab("Distance from gene start site (binned to 100 bp)") + ylab("Average Conservation Score") +
  scale_x_continuous(breaks = seq(-1500,10000,1500))
cat("\n")
ggsave("CpGdist.svg")
```

@@

cat("\n")
# Fig 4A: Example genes CpG site distributions
plotGeneScore("TUBA1A",0.2)
#ggsave("genescore_TUBA1A.svg")
cat("\n")
plotGeneScore("TTN",0.26)
#ggsave("genescore_TTN.svg")
cat("\n")
plotGeneScore("HLA-A",0.18)
#ggsave("genescore_HLA-A.svg")
cat("\n")
plotGeneScore("APC",0.12) + theme(aspect.ratio = 1)
#ggsave("genescore_APC.svg")
cat("\n")
plotGeneScore("KRAS",0.18) + theme(aspect.ratio = 1)
#ggsave("genescore_KRAS.svg")

@@

# CRC gene database 1: COSMIC
cancergenes1 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
# CRC gene database 2: AGO
cancergenes2 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
essentialgenes <- read_delim("/Users/kevinmurgas/Documents/Data+ project/Manuscript/Bioinformatics Revision/core-essential-genes-sym_HGNCID.txt", delim = "\t", col_names = F)

@@

# plot violins of: all genes, COSMIC, AGO
perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n)) %>% mutate(score_median = 2*score_median)
genes_dbs <- rbind(perm_temp %>% mutate(group="all genes"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% mutate(group="COSMIC"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (essentialgenes %>% pull(X1))) %>% mutate(group="DEPMAP")
              )
# Fig 4C: plot distribution
pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all genes", "AGO", "COSMIC", "DEPMAP"))) %>%
  #ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) +
  ggplot() + geom_boxplot(aes(x=score_median,y=group,fill=group), outlier.shape = NA) +
  scale_x_continuous(limits=c(-5,7), breaks = c(-4, -2,0, 2, 4, 6)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
# # add hashmarks for each gene (doesn't look great)
# marks <- genes_dbs %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
# marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
# pDB + geom_segment(data=marks2, aes(x=x, xend=x, y=y, yend=y+0.2), color="red") + geom_segment(data=marks, aes(x=x, xend=x, y=y, yend=y+0.1), color="black", alpha=0.5)
pDB
# ggsave("database_violins.svg")
ggsave("database_boxplot_depmap.svg")

genes_dbs %>% group_by(group) %>% summarize(mean = mean(score_median))

# stats
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="AGO"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="DEPMAP"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
```

\newpage
\subsection{Oncogenes vs Tumor suppressors}
Based on list of top 20 genes mutated in CRC, manually labelled as oncogene or tumor-suppressor:

| Oncogene | Tumor Suppressor |
|----------|------------------|
| KRAS     | TP53             |
| MYC      | APC              |
| BRAF     | KLF4             |
| EGFR     | FBXW7            |
| CDK8     | LRP1B            |
| PIK3CA   | PTEN             |
| NRAS     | FAT4             |
| GNAS     | ARID1A           |
| SRC      | AMER1            |
| CTNNB1   | SMAD4            |

```{r Onco_vs_TSG, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA, fig.fullwidth=TRUE, dev = "png"}

plotGeneScore <- function(geneStr,norm) {
  cat("\n")
  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
    scale_x_continuous(limits=c(-5,5), breaks=c(-4,-2,0,2,4)) +
    scale_y_continuous(limits=c(0,0.03),breaks=c(0,0.01,0.02,0.03)) +
    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
}

cat("\nOncogenes:")
plotGeneScore("KRAS",0.25)
plotGeneScore("MYC",0.21)
plotGeneScore("BRAF",0.5)
plotGeneScore("EGFR",0.5)
plotGeneScore("CDK8",0.5)
plotGeneScore("PIK3CA",0.5)
plotGeneScore("NRAS",0.5)
plotGeneScore("GNAS",0.5)
plotGeneScore("SRC",0.5)
plotGeneScore("CTNNB1",0.5)

cat("\nTumor Suppressors:")
plotGeneScore("TP53",0.25)
plotGeneScore("APC",0.21)
plotGeneScore("KLF4",0.5)
plotGeneScore("FBXW7",0.5)
plotGeneScore("LRP1B",0.5)
plotGeneScore("PTEN",0.5)
plotGeneScore("FAT4",0.5)
plotGeneScore("ARID1A",0.5)
plotGeneScore("AMER1",0.5)
plotGeneScore("SMAD4",0.5)
